Compositional Certified Resource Bounds

### 2014-07-01 paper/main.tex
updates to the disussion of the soundness of the logic

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as
\begin{align*}
  \forall \state \, \cost \, m \, \cost' .\,&
  (m \le n \land P(\state) \le \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
  &\implies \cost' \ge 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used in the
soundness proof for the function call and loop cases.  The constraint
$\cost' \geq 0$ imposed by the definition ensures that the program does
not stop because of a resource error (recall that a negative resource
counter on the right-hand side is a resource failure).  However, it
does not rule out memory safety errors or assertion failures, this is
because our logic does not prove any safety or correctness theorems
but only focuses on resource usage.

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcome of a program
statement.  We denote it by $\safeK n B R Q \cont$ and define
it to be
\begin{align*}
& \safe n B {\code{break}} \cont \\
\land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
\land& \safe n Q {\code{skip}} \cont.
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as follows
\begin{align*}
&\valid n B R P S Q := \\
&\qquad \forall m \, \cont \, x .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
&\qquad \implies \safe m {P+x} S \cont.
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic, this refinement is necessary to have a stronger
induction hypothesis available during the proof.

Using the semantic validity of triples we can define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.  A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity is constrained by
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two contraints force
$\delta$ to be exactly one in the theorem statement.

If $S$ is a complete program and $\Delta$ is empty, by
expanding the definitions we see that $\Delta$ is valid
for every $n$ and that $\Kstop$ is safe for every $n$, so
we get
$$
  \forall n .\, \safe n P S \Kstop.
$$
This means that from any starting state $\state$, $P(\state)$
provides enough resource for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.

>>>>

We already gave an intuition of the meaning of judgements
derived in the logic.  To make it formal, we define the
\emph{resource safety} $\safe n P S \cont$ of an assertion
and a program configuration as
\begin{align*}
  \forall \state \, \cost \, m \, \cost' .\,&
  (m \le n \land P(\state) \le \cost \land
    (\state, S, \cont, \cost) \smallstep{M}^m (\_, \_, \_, \cost')) \\
  &\implies \cost' \ge 0.
\end{align*}
This predicate is step indexed by an integer $n$ that is used in the
soundness proof for the function-call and loop cases.  The constraint
$\cost' \geq 0$ imposed by the definition ensures that the program does
not stop because of a resource error (recall that a negative resource
counter on the right-hand side is a resource failure).  However, it
does not rule out memory safety errors or assertion failures. This is
because our logic does not prove any safety or correctness theorems
but only focuses on resource usage.
%
An interesting detail in the definition is the natural number $m$.  We
simply use it to ensure that $\safe {n+1} P S \cont \implies \safe n P S
\cont $.  This would not be the case if we replaced all occurrences of
$m$ by $n$ in the definition of validity.

The resource safety of a continuation $\cont$ is defined using three
assertions, one for each of the possible outcomes of a program
statement.  We define it as follows.
\begin{align*}
\safeK n B R Q \cont & := & & \safe n B {\code{break}} \cont \\
& & \land& \safe n {\lambda \state.\,R\,(\state(x))} {\code{return}~x} \cont \\
& & \land& \safe n Q {\code{skip}} \cont
\end{align*}
%
We can now define the \emph{semantic validity} of a judgement $B; R
\vdash \htriple P S Q$ of the quantitative logic without function
context as $\valid n B R P S Q :=$
\begin{align*}
& \forall m \, \cont \, x \geq 0 .\, (m \le n \land \safeK m {B+x} {R+x} {Q+x} \cont) \\
& \implies \safe m {P+x} S \cont \; .
\end{align*}
Note how the validity of a triple embeds the frame rule of
our logic. This refinement is necessary to have a stronger
induction hypothesis available during the proof.
%
We again need to add the auxiliary $m$ to ensure that $\valid {n{+}1} B R P
S Q$ implies $\valid n B R P S Q$.

Using the semantic validity of triples we can define the validity
of a function context $\Delta$, written $\validC n \Delta$, as
\begin{align*}
  &\forall f .\, \Delta(f) =
    \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v)
    \implies \\
  &\qquad \forall z \, \vec v .\,
  \valid n \bot {Q_f \, z} {P_f \, z \, \vec v} {S_f} \bot ,
\end{align*}
where $S_f$ is the body of the function $f$.  A full
judgement that mentions a non-empty function context
$\Delta$ is in fact a guarded statement: it makes
assumptions on some functions' behavior.  The predicate
$\validC n \Delta$ gives the precise meaning of the
assumptions made.  It is also step-indexed to prove the
soundness of the {\sc L:Extend} rule by induction.
%
\begin{theorem}[Soundness of the logic]
  If $\Delta; B; R \vdash \htriple P S Q$ is derivable, then
  $$
    \forall n .\, \validC n \Delta
      \implies \valid {n {+} 1} B R P S Q.
  $$
\end{theorem}
%
\noindent
The difference $\delta = 1$ between the index in the triple
validity and the one in context validity arises from
the soundness proofs of {\sc L:Call} and {\sc L:Extend}.  For
{\sc L:Call}, the language semantics makes one step and
proceeds with the function body, so we must have
$\delta \le 1$ to use the assumptions in $\Delta$.
For {\sc L:Extend}, we have to show that $\Delta \cup \Delta'$
is a valid context for $n$ steps.  The induction hypothesis
in that case says that if $\Delta \cup \Delta'$ is valid
for $m$ steps, $\Delta'$ is valid for $m+\delta$ steps.
So if we want to solve this goal by induction, it is
necessary that $\delta \ge 1$.  These two contraints force
$\delta$ to be exactly one in the theorem statement.

If $S$ is a complete program and $\Delta$ is empty, by
expanding the definitions we see that $\Delta$ is valid
for every $n$ and that $\Kstop$ is safe for every $n$, so
we get
$$
\Delta; B; R \vdash \htriple P S Q \implies   \forall n .\, \safe n P S \Kstop.
$$
This means that from any starting state $\state$, $P(\state)$
provides enough resources for any run of the program $S$.  This
setting is actually the main use case of the previous theorem
which is stated as a stronger result to allow a proof by
induction.
